Temporal ordering and registration of images in studies of developmental dynamics

### 2015-02-04 image_analysis_paper_development2/manuscript.tex
after yannis comments; before stas comments

\customlabel{fig:DV_view}{S1}
\customlabel{fig:1d_example}{S2}
\customlabel{fig:eigenvalues_drosophila_live}{S3}
\customlabel{fig:eigenvalues_zebrafish}{S4}
\customlabel{fig:eigenvalues_drosophila_fixed}{S5}
\customlabel{fig:eigenvalues_wing_disc}{S6}

@@

\author{Carmeline~J.~Dsilva \thanks{Department of Chemical and Biological Engineering, Princeton University, Princeton, New Jersey, USA},
Bomyi~Lim \footnotemark[1],
Hang~Lu \thanks{School of Chemical and Biomolecular Engineering, Georgia Institute of Technology, Atlanta, Georgia, USA},
Amit~Singer \thanks{Department of Mathematics, Princeton University, Princeton, New Jersey, USA} \thanks{Program in Applied and Computational Mathematics, Princeton University, Princeton, New Jersey, USA},
Stanislav~Y.~Shvartsman \footnotemark[1] \thanks{Lewis-Sigler Institute for Integrative Genomics, Princeton University, Princeton, New Jersey, USA}, \\
and Ioannis~G.~Kevrekidis \footnotemark[1] \footnotemark[4]}

@@

We present a robust algorithmic approach to registration and temporal ordering.
%
In contrast to a number of previous methodologies \citep{zitova2003image, rowley1998rotation, hajnal2010medical, greenspan1994rotation, zhao2003face}, our methodology does not rely on the {\em a priori} knowledge of spatial landmarks for registration or markers of developmental progression for temporal ordering.
%
Our approach is based on a manifold learning algorithm (vector diffusion maps \citep{singer2012vector}) which simultaneously addresses the problems of registration and temporal ordering. 
%
This algorithm is one of several nonlinear dimensionality reduction techniques that have been developed over the past decade \citep{Belkin2003, coifman2005geometric, coifman2006geometric, tenenbaum2000global, roweis2000nonlinear}, for
applications ranging from analysis of cryo-electron microscopy (cryo-EM) images of individual molecules  \citep{zhao2014rotationally, singer2011viewing} to face recognition \citep{lafon2006data} and classification of CT scans \citep{fernandez2014diffusion}.
%
Here, the vector diffusion maps algorithm is adapted for the analysis of images of developing embryos in studies of developmental dynamics, with the main objective of revealing stereotypic developmental trajectories from fixed images.
%
To illustrate our approach, we analyze four data sets common in developmental biology. 
%
Our first two data sets come from live imaging studies of {\em Drosophila} and zebrafish embryogenesis.
%
In both of these examples, the correct rotational orientation and temporal order are independently known, and these data sets will be used to validate our approach.
%from a study of {\it Drosophila} embryogenesis, one of the best experimental models for studies of development \citep{jaeger2012drosophila}.
%
Our third data set consists of images from fixed {\em Drosophila} embryos where the correct orientation and order is unknown; here, we will show how the algorithm can help uncover developmental dynamics which are not readily apparent. 
%
Our final data set consists of z-stacks of wing discs, which we will use to illustrate how our methods extend to three-dimensional imaging. 
%
We will also show how to compute an average trajectory from a set of registered and ordered fixed images to remove noise due to intersample variability and obtain a smooth description of the underlying developmental dynamics.

@@

Vector diffusion maps \citep{singer2012vector} is a manifold learning
technique developed for data sets which contain two sources of variability:
geometric symmetries (such as rotations of the images) which one would like to factor out,
and ``additional" directions of variability (such as temporal dynamics) which one would like to uncover.
%
Vector diffusion maps combine two algorithms, {\em angular synchronization} \citep{singer2011angular} for image registration and {\em diffusion maps} \citep{coifman2005geometric} for extracting intrinsic low-dimensional structure in data, into a single computation.
%
We will use the algorithm to register images of developing embryos with respect to rotations, as well as uncover the main direction of variability {\it after} removing rotational symmetries.
%
Although in general, images may contain variations due to rotations, translations, and scaling, we will remove relevant translations and/or scaling via relatively simple preprocessing, and focus only on factoring out rotations using the vector diffusion maps algorithm.
%
In the case that all relevant symmetries can be removed with straightforward preprocessing procedures, our algorithms can also be used simply to extract the main direction of variability within the imaging data set. 
%
We assume that the main direction of remaining variability in these images is parameterized by the developmental time of each embryo.
%
As a consequence, uncovering this direction should reveal the underlying developmental dynamics.

\begin{figure}[t]
\includegraphics{fig2}
\caption{Schematic illustrating angular synchronization and diffusion maps. (A) Set of vectors, each in a different orientation. The pairwise alignment angles are indicated. (B) The vectors from A, each rotated about their midpoint so that the set is globally aligned. Note that the chosen rotation angles are consistent with the pairwise alignments in A: the differences between a pair of angles in B is the same as the pairwise angle in A. (C) Data points (in black) which lie on a one-dimensional nonlinear curve in two dimensions. Each pair of points is connected by an edge, and the edge weight is related to the Euclidean distance between the points through a Gaussian kernel (see \SI), so that close data points are connected by darker (``stronger'') edges. (D) The data in C, colored by the first (non-trivial) eigenvector from the diffusion map computational procedure. The color intensity is monotonic with the perceived curve arclength, thus parameterizing the curve.}
\label{fig:schematics}
\end{figure}

Angular synchronization uses pairwise alignment information to register a set of images in a globally consistent way.
%
A schematic illustration of angular synchronization is shown in \fig~\ref{fig:schematics}A, where each image is represented as a vector, and the goal is to align the entire set of vectors given pairwise alignment measurements.
%
We first compute the angles needed to align pairs of vectors (or images), which in general requires no template function \citep{ahuja2007template}.
%
In this work, we aligned pairs of images with respect to rotations by exhaustively searching over a discretized space of rotation angles. 
%
However, in other applications, pairwise alignments can be computed by aligning appropriate image landmarks or features. 
%
When the data are noisy, these pairwise measurements may by inaccurate, and so we would like to utilize {\em all} pairwise measurements to align the set of images robustly. 
%
Using the alignment angles between all pairs of vectors, angular synchronization finds the set of rotation angles (one angle for each vector) that is most consistent with {\it all} pairwise measurements (see \SI); this is illustrated in \fig~\ref{fig:schematics}B.
%
In this schematic, registration via angular synchronization is trivial, as the pairwise measurements contain no noise.
%
However, the algorithm can successfully register data sets even when many of the pairwise measurements are inaccurate \citep{singer2011angular}.

After removing variability due to rotations, the developmental dynamics may be revealed by ordering the data along the one-dimensional manifold that parameterizes most of the remaining variability in the data.
%
Such a manifold can be discovered using diffusion maps \citep{coifman2005geometric}, a nonlinear dimensionality reduction technique that uncovers an intrinsic parameterization of data that lies on a low-dimensional manifold in high-dimensional space.
%
The idea is illustrated in \fig~\ref{fig:schematics}C, where the data are two-dimensional points which lie on a one-dimensional nonlinear curve.
%
We use {\it local} information about the data to find a parameterization which respects the underlying manifold geometry, so that points which are close in high-dimensional space (e.g., images which look similar) are close in our parameterization.
%
This idea of locality is denoted by the color of the edges in \fig~\ref{fig:schematics}C:
data points which are close are connected by dark edges, and clearly, the dark edges are more ``informative" about the low-dimensional structure of the data.
%
The color in \fig~\ref{fig:schematics}D depicts the one-dimensional parameterization or ordering of the data that we can detect visually.
%
In our working examples, each data point will be of much higher dimension (e.g., a pixelated image or three-dimensional voxel data), and so we cannot extract this low-dimensional structure visually.
%
Instead, we will use diffusion maps to automatically uncover a parameterization of our high-dimensional data from the eigenvectors of the appropriate matrix (see \SI).
%
Furthermore, the corresponding eigenvalues will allow us to confirm our assumption that our data lie on a one-dimensional manifold. 
%
Diffusion maps generalizes directly from one-dimensional nonlinear curves to higher-dimensional manifolds.

@@

\raisebox{4.25cm}{\textsf{C}}
\includegraphics[width=5cm]{drosophila_live_imaging_angle_corr}
%
%\hfill
%
\raisebox{4.25cm}{\textsf{D}}
\includegraphics[width=5cm]{drosophila_live_imaging_rank_corr}
%
%\hfill
%
\raisebox{4.25cm}{\textsf{E}}
%
\raisebox{2.5cm}{
{ \textsf{ 
\begin{tabular}{ |c | c| c| }
\hline
\parbox[c]{1.5cm}{ \centering Movie Index} & \parbox[c]{1.5cm}{ \vspace{.5\baselineskip} \centering Average Angle Error \vspace{.5\baselineskip}} & \parbox[c]{1.5cm}{ \centering Rank Correlation Coefficient} \\
\hline
1 & 8.37$^{\circ}$ & 0.9989 \\ 
2 & 4.98$^{\circ}$ & 0.9994 \\ 
3 & 6.47$^{\circ}$ & 0.9795 \\ 
4 & 7.03$^{\circ}$ & 0.9953 \\ 
5 & 3.04$^{\circ}$ & 0.9959 \\ 
 \hline
\end{tabular}}}}
\caption{Method validation using live imaging of \textit{Drosophila} embryos. (A) Selected images from a live imaging study of a \textit{Drosophilla} embryo during gastrulation. Each frame is in an arbitrary rotational orientation, and the order of the frames has been shuffled. (B) Images from A after registration and ordering using vector diffusion maps. The dorsal side of each embryo now appears at the top of each image, and the ventral side appears at the bottom. (C) The correlation between the recovered rotation angle (using vector diffusion maps) and the true rotation angle. The average absolute error in the recovered angles is $8.37 ^\circ$. (D) The correlation between the recovered rank (using vector diffusion maps) and the true rank. The rank correlation coefficient is $0.9989$. (E) The average error in the recovered angle and the rank correlation coefficient for $5$ different live imaging studies. }
\label{fig:drosophila_live_imaging}
\end{figure*}

To validate the proposed approach, we applied our algorithm to a data set where the true temporal order and rotational orientation of the images were known {\em a priori}.
%
This data set was obtained through live imaging near the posterior pole of a vertically oriented {\it Drosophila} embryo during the twenty minutes spanning the late stages of cellularization through early gastrulation.
%
During this time window, the ventral furrow is formed, where the ventral side buckles towards the center of the embryo, internalizing the future muscle cells and forming a characteristic ``omega'' shape.
%
Germband extension then causes cells from the ventral side to move towards the posterior pole of the embryo, and then wrap around to the dorsal side \citep{leptin2005gastrulation}.
%
At the end of this process, cells which were originally on the ventral and posterior side of the embryo find themselves on the dorsal side, causing a similar ``omega'' to appear on the dorsal side.

@@

We used vector diffusion maps to register and order the images. 
%
\fig~\ref{fig:drosophila_live_imaging}B shows the images from \fig~\ref{fig:drosophila_live_imaging}A, now registered and ordered; the real time for each frame is also indicated.
%
With a small number of exceptions, the recovered ordering is consistent with the real time dynamics. 
%
\fig~\ref{fig:drosophila_live_imaging}C and \fig~\ref{fig:drosophila_live_imaging}D  show the correlations between the recovered and true angles and ranks, respectively, for the entire data set. 
%
Both the angles and the ranks are recovered with a high degree of accuracy.

@@

\subsubsection*{Zebrafish gastrulation}

@@

\vspace{0cm}
\raisebox{2.5cm}{\textsf{C}}
%
\includegraphics[width=3.25cm]{zebrafish_rank_corr}
\end{minipage}
\caption{Method validation using live imaging of a zebrafish embryo. (A) Selected images from a flip book movie of zebrafish embryogenesis. Each frame is in an arbitrary rotational orientation, and the order of the frames has been shuffled. (B) Images from A after registration and ordering using vector diffusion maps. The real time of each frame is also indicated. (C) Correlation between the rank recovered using vector diffusion maps and the true rank.  The rank correlation coefficient is $0.9954$. The larger errors in the recovered ranks towards the beginning of the trajectory are due to the rather static cell movement within that time window. }
\label{fig:zebrafish}
\end{figure*}

To illustrate the broad applicability of our approach, we also applied our algorithm to a time-lapse movie  of zebrafish embryogenesis. 
%
We used a publicly available flipbook movie of zebrafish embryogenesis, available at \texttt{https://zfin.org/zf\_info/movies/Zebrafish.mov} \citep{karlstrom1996flipbook}.  
%
Taken with a differential interference contrast (DIC) microscope, the movie records the first $17$~hours of zebrafish development, from a single cell stage to a 16-somite stage. 
%
We selected $120$ consecutive frames from this movie which capture 5.5~hours of epiboly (3.5--9~hours after fertilization). 
%
At the start of the time window, cells have divided 10--11 times and are accumulated in a cell mass above the yolk. 
%
The cell mass is then compressed and the animal-vegetal axis of the embryo (vertical axis in the \fig~\ref{fig:zebrafish}) shortens to form a spherical embryo shape by the end of the fourth hour of development. 
%
Then, the yolk syncytial layer, which forms the boundary between the yolk and the cell mass, moves upward, forming a dome-shaped structure.  
%
During this stage, the cells rearrange to form a cell layer with uniform thickness (about four cells thick). 
%
With time, this cell layer then spreads over across the yolk and expands toward the vegetal pole. 
%
At the end of epiboly, the blastoderm completely engulfs the yolk.

As in the example of {\em Drosophila} embryo live imaging, the frames were randomly rotated and shuffled (\fig~\ref{fig:zebrafish}A). 
%
We then used vector diffusion maps to register and order the frames. 
%
The results are shown in \fig~\ref{fig:zebrafish}B. 
%
The recovered rotations and order are consistent with the expected developmental dynamics, as shown in the correlations between the recovered and true ranks (\fig~\ref{fig:zebrafish}C). 
%
Quantitatively, the rank correlation coefficient for this data set is 0.9954, and the average error in the recovered angle is $4.14^\circ$. 
%
Some errors in ordering images of the early embryo result from rather static cell movement during the early developmental stage where cells divide and accumulate above the yolk. 
%
During epiboly, cell movement is more dynamic and the recovered ordering is more consistent with the real dynamics.

In summary, we showed that our algorithm is not limited to a specific set of data. 
%
The algorithm works for different tissue types ({\em Drosophila} and zebrafish), and also for images taken with different imaging methods (fluorescent and phase contrast images).
%
Provided there exist significant dynamics within the data set and that the developmental trajectory is well-sampled, the developmental dynamics can be recovered.

@@

\raisebox{0.5cm}{\textsf{C}}
\includegraphics[width=12.25cm]{drosophila_fixed_images_average}
\end{minipage}
%
\hfill
%
\raisebox{3.25cm}{\textsf{D}}
\includegraphics[width=4.25cm]{drosophila_fixed_images_rank_corr}
%
\caption{Analyzing images of fixed \textit{Drosophila} embryos. {(A)} Images of \textit{ Drosophila} embryos, stained for nuclei (gray), Twi (green), and dpERK (red). Each image is of a different embryo arrested at a different developmental time and in a different rotational orientation. {(B)} Data from {\it A}, registered and ordered using vector diffusion maps. The expert rank for each image is indicated. {(C)} A representative ``developmental trajectory'' obtained from local averaging of the entire set of registered and ordered images (see \SI). {(D)} Correlation between the image ranks calculated from the vector diffusion maps algorithm and the ranks obtained from ordering by an expert. The rank correlation coefficient is $0.9716$. }
\label{fig:drosophila_fixed_images}
\end{figure*}

We have analyzed how our algorithm performs on two model data sets where all images come from a single embryo. 
%
In practice, we are interested in cases where image comes from a different embryo, and the largest source of noise in the considered data set arises from embryo-to-embryo variability.
%
To demonstrate that our methods are robust to such noise sources,
we constructed a synthetic time course data set by selecting a random image from one of five {\em Drosophila} live imaging data sets (those data sets used in \fig~\ref{fig:drosophila_live_imaging}) at each time point.
%
The resulting data set is spatially unregistered, scrambled in time, and reflects embryo-to-embryo variability. 
%
The median rank correlation coefficient when ordering such a synthetic time course using our methodology was 0.77, indicating that the algorithm can accurately recover the temporal order even under noisy conditions.

@@

\fig~\ref{fig:drosophila_fixed_images}B shows the selected images in \fig~\ref{fig:drosophila_fixed_images}A, now registered and ordered using vector diffusion maps \citep{singer2012vector}.
%
However, because of interembryo variability, these selected images are not entirely representative of the developmental dynamics across the entire data set.
%
To highlight the developmental dynamics revealed by vector diffusion maps, we computed an average developmental trajectory across the set of registered and ordered images (see \SI). 
%
Each snapshot in the average trajectory is the (weighted) average of a group of successive images from the registered and ordered data set.
%
Averaging successive images removes (some) of the interembryo variability, so that sequential snapshots of this averaged trajectory, shown in \fig~\ref{fig:drosophila_fixed_images}C, serve as a summary of the stereotypic developmental dynamics.

From this average trajectory, we can now easily see the developmental progression, which is consistent with the known dynamics: 
dpERK first appears as two lateral peaks at the ventral side of the embryo; a third dpERK peak then appears at the dorsal side of the embryo.
%
During mesoderm invagination, the two ventral dpERK peaks merge together, eventually forming, together with Twi, the ``omega'' shape.
%
The dorsal dpERK peak then disappears during germband extension as cells from the ventral side wrap around to the dorsal side.
%
At the end of this process, similar ``omegas'' formed by Twi and dpERK appear on the dorsal side of the embryo; these patterns are most readily seen in the last image of \fig~\ref{fig:drosophila_fixed_images}C.
%
Thus, vector diffusion maps can accomplish the tasks presented in the caricature in \fig~\ref{fig:fish}, even in the absence of information about image landmarks (e.g., fins) and without {\it a priori} knowledge of developmental features (e.g., correlation of age with body size).
%

>>>>

\customlabel{fig:DV_view}{S1}
\customlabel{fig:1d_example}{S2}
\customlabel{fig:eigenvalues_drosophila_live}{S3}
\customlabel{fig:eigenvalues_zebrafish}{S4}
\customlabel{fig:eigenvalues_drosophila_fixed}{S5}
\customlabel{fig:eigenvalues_wing_disc}{S6}
\customlabel{fig:PCA_compare}{S7}

@@

\author{Carmeline~J.~Dsilva \thanks{Department of Chemical and Biological Engineering, Princeton University, Princeton, New Jersey, USA},
Bomyi~Lim \footnotemark[1],
Ioannis~G.~Kevrekidis \footnotemark[1] \thanks{Program in Applied and Computational Mathematics, Princeton University, Princeton, New Jersey, USA} ,
Hang~Lu \thanks{School of Chemical and Biomolecular Engineering, Georgia Institute of Technology, Atlanta, Georgia, USA},
Amit~Singer \thanks{Department of Mathematics, Princeton University, Princeton, New Jersey, USA} \footnotemark[2], \\ and
Stanislav~Y.~Shvartsman \footnotemark[1] \thanks{Lewis-Sigler Institute for Integrative Genomics, Princeton University, Princeton, New Jersey, USA}}

@@

We present a robust algorithmic approach to simultaneous registration and temporal ordering.
%
In contrast to a number of previous methodologies \citep{zitova2003image, rowley1998rotation, hajnal2010medical, greenspan1994rotation, zhao2003face, dubuis2013accurate}, our methodology does not rely on the {\em a priori} knowledge of spatial landmarks for registration or markers of developmental progression.
%
The approach is based on a manifold learning algorithm (vector diffusion maps \citep{singer2012vector}) which simultaneously addresses the problems of registration and temporal ordering. 
%
This algorithm is one of several nonlinear dimensionality reduction techniques that have been developed over the past decade \citep{Belkin2003, coifman2005geometric, coifman2006geometric, tenenbaum2000global, roweis2000nonlinear}, for
applications ranging from analysis of cryo-electron microscopy (cryo-EM) images of individual molecules  \citep{zhao2014rotationally, singer2011viewing} to face recognition \citep{lafon2006data} and classification of CT scans \citep{fernandez2014diffusion}.
%
Here, the vector diffusion maps algorithm is adapted for the analysis of images of developing tissues in studies of developmental dynamics, with the main objective of revealing stereotypic developmental trajectories from fixed images.
%
To illustrate our approach, we analyze four data sets common in developmental biology. 
%
Our first two data sets come from live imaging studies of {\em Drosophila} and zebrafish embryogenesis.
%
In both of these examples, the correct rotational orientation and temporal order are independently known, and these data sets will be used to validate our approach.
%from a study of {\it Drosophila} embryogenesis, one of the best experimental models for studies of development \citep{jaeger2012drosophila}.
%
Our third data set consists of images from fixed {\em Drosophila} embryos where the correct orientation and order is unknown; here, we will show how the algorithm can help uncover developmental dynamics which are not readily apparent. 
%
Our final data set consists of z-stacks of wing discs, which we will use to illustrate how our methods can be used to analyze specific types of three-dimensional imaging data. 
%
We will also show how to compute an average trajectory from a set of registered and ordered fixed images to remove noise due to intersample variability and obtain a smooth description of the underlying developmental dynamics.

@@

Vector diffusion maps \citep{singer2012vector} is a manifold learning
technique developed for data sets which contain two sources of variability:
geometric symmetries (such as rotations of the images) which one would like to factor out,
and ``additional" directions of variability (such as temporal dynamics) which one would like to uncover.
%
Vector diffusion maps combine two algorithms, {\em angular synchronization} \citep{singer2011angular} for image registration and {\em diffusion maps} \citep{coifman2005geometric} for extracting intrinsic low-dimensional structure in data, into a single computation.
%
We will use the algorithm to register images of developing tissues with respect to rotations, as well as uncover the main direction of variability {\it after} removing rotational symmetries.
%
Although in general, images may contain variations due to rotations, translations, and scaling, we will remove the relevant translational and/or scaling symmetries via relatively simple image preprocessing, and focus here only on factoring out rotations using the vector diffusion maps algorithm.
%
In the case that {\em all} relevant symmetries can be removed with straightforward preprocessing procedures, our algorithms can also be used simply to extract the main direction of variability within the imaging data set. 
%
We assume that this main direction of remaining variability in these images is parameterized by the developmental time of each embryo.
%
As a consequence, uncovering this direction should reveal the underlying developmental dynamics.

\begin{figure}[t]
\includegraphics{fig2}
\caption{Schematic illustrating angular synchronization and diffusion maps. (A) Set of vectors, each in a different orientation. The pairwise alignment angles are indicated. (B) The vectors from A, each rotated about their midpoint so that the set is globally aligned. Note that the chosen rotation angles are consistent with the pairwise alignments in A: the difference between a pair of angles in B is the same as the pairwise angle in A. (C) Data points (in black) which lie on a one-dimensional nonlinear curve in two dimensions. Each pair of points is connected by an edge, and the edge weight is related to the Euclidean distance between the points through a Gaussian kernel (see \SI), so that pairs of data point which are close are connected by darker (``stronger'') edges. (D) The data in C, colored by the first (non-trivial) eigenvector from the diffusion map computational procedure. The color intensity is monotonic with the perceived curve arclength, thus parameterizing the curve.}
\label{fig:schematics}
\end{figure}

Angular synchronization uses pairwise alignment information to register a set of images in a globally consistent way.
%
A schematic illustration of angular synchronization is shown in \fig~\ref{fig:schematics}A, where each image is represented as a vector, and the goal is to align the entire set of vectors given pairwise alignment measurements.
%
We first compute the angles needed to align pairs of vectors (or images), which in general requires no notion of a template function \citep{ahuja2007template}.
%
In this work, we aligned pairs of images with respect to rotations by exhaustively searching over a discretized space of rotation angles. 
%
However, pairwise alignments can be computed by aligning appropriate image landmarks or features \citep{ian1998statistical}. 
%
When the data are noisy, these pairwise measurements may by inaccurate, and so we utilize {\em all} pairwise measurements to align the set of images robustly. 
%
Using the alignment angles between all pairs of vectors, angular synchronization finds the set of rotation angles (one angle for each vector) that is most consistent with {\it all} pairwise measurements (see \SI); this is illustrated in \fig~\ref{fig:schematics}B.
%
In this schematic, registration via angular synchronization is trivial, as the pairwise measurements contain no noise.
%
However, the algorithm can register data sets even when many of the pairwise measurements are inaccurate \citep{singer2011angular}.

After removing variability due to rotations, the developmental dynamics may be revealed by ordering the data along the one-dimensional curve that parameterizes most of the remaining variability in the data.
%
Such a curve can be discovered using diffusion maps \citep{coifman2005geometric}, a nonlinear dimensionality reduction technique that reveals a parameterization of data that lies on a low-dimensional manifold in high-dimensional space.
%
The idea is illustrated in \fig~\ref{fig:schematics}C, where the data are two-dimensional points which lie on a one-dimensional (nonlinear) curve.
%
We use {\it local} information about the data to find a parameterization which respects the underlying manifold geometry, so that points which are close in high-dimensional space (e.g., images which look similar) are close in our parameterization.
%
This idea of locality is denoted by the color of the edges in \fig~\ref{fig:schematics}C:
data points which are close are connected by dark edges, and clearly, the dark edges are more ``informative" about the low-dimensional structure of the data.
%
The color in \fig~\ref{fig:schematics}D depicts the one-dimensional parameterization or ordering of the data that we can detect visually.
%
A detailed example of using vector diffusion maps to register and order synthetic data is given in \fig~\ref{fig:1d_example}, and a step-by-step tutorial of the diffusion maps implementation is included in the available software. 
%
In our working examples, each data point will be of much higher dimension (e.g., a pixelated image or three-dimensional voxel data), and so we cannot extract this low-dimensional structure visually.
%
Instead, we will use diffusion maps which automatically uncovers a parameterization of our high-dimensional data from the eigenvectors of the appropriate matrix (see \SI).
%
Furthermore, the corresponding eigenvalues will allow us to confirm our assumption that our data lie on a one-dimensional manifold (see \fig~\ref{fig:eigenvalues_drosophila_live}--\ref{fig:eigenvalues_wing_disc}). 
%
Diffusion maps generalizes directly from one-dimensional nonlinear curves to higher-dimensional manifolds.

@@

\raisebox{4.25cm}{\textsf{C}}
\includegraphics[width=5cm]{drosophila_live_imaging_angle_corr}
%
%\hfill
%
\raisebox{4.25cm}{\textsf{D}}
\includegraphics[width=5cm]{drosophila_live_imaging_rank_corr}
%
%\hfill
%
\raisebox{4.25cm}{\textsf{E}}
%
\raisebox{2.5cm}{
{ \textsf{ 
\begin{tabular}{ |c | c| c| }
\hline
\parbox[c]{1.5cm}{ \centering Movie Index} & \parbox[c]{1.5cm}{ \vspace{.5\baselineskip} \centering Average Angle Error \vspace{.5\baselineskip}} & \parbox[c]{1.5cm}{ \centering Rank Correlation Coefficient} \\
\hline
1 & 8.37$^{\circ}$ & 0.9989 \\ 
2 & 4.98$^{\circ}$ & 0.9994 \\ 
3 & 6.47$^{\circ}$ & 0.9795 \\ 
4 & 7.03$^{\circ}$ & 0.9953 \\ 
5 & 3.04$^{\circ}$ & 0.9959 \\ 
 \hline
\end{tabular}}}}
\caption{Method validation using live imaging of \textit{Drosophila} embryos. (A) Selected images from a live imaging study of a \textit{Drosophila} embryo during gastrulation. Each frame is in an arbitrary rotational orientation, and the order of the frames has been shuffled. (B) Images from A after registration and ordering using vector diffusion maps. The dorsal side of each embryo now appears at the top of each image, and the ventral side appears at the bottom. (C) The correlation between the recovered rotation angle (using vector diffusion maps) and the true rotation angle. The average absolute error in the recovered angles is $8.37 ^\circ$. (D) The correlation between the recovered rank (using vector diffusion maps) and the true rank. The rank correlation coefficient is $0.9989$. (E) The average error in the recovered angle and the rank correlation coefficient for $5$ independent live imaging studies. }
\label{fig:drosophila_live_imaging}
\end{figure*}

To validate the proposed approach, we first applied our algorithm to a data set where the true temporal order and rotational orientation of the images were known {\em a priori}.
%
This data set was obtained through live imaging near the posterior pole of a vertically oriented {\it Drosophila} embryo during the twenty minutes spanning the late stages of cellularization through early gastrulation.
%
During this time window, the ventral furrow is formed, where the ventral side buckles towards the center of the embryo, internalizing the future muscle cells and forming a characteristic ``omega'' shape.
%
Germband extension then causes cells from the ventral side to move towards the posterior pole of the embryo, and then wrap around to the dorsal side \citep{leptin2005gastrulation}.
%
At the end of this process, cells which were originally on the ventral and posterior side of the embryo find themselves on the dorsal side, causing a similar ``omega'' to appear on the dorsal side.

@@

We used vector diffusion maps to register and order the images. 
%
\fig~\ref{fig:drosophila_live_imaging}B shows the images from \fig~\ref{fig:drosophila_live_imaging}A, now registered and ordered; the real time for each frame is also indicated.
%
With a small number of exceptions, the recovered ordering is consistent with the real time dynamics. 
%
\fig~\ref{fig:drosophila_live_imaging}C and \fig~\ref{fig:drosophila_live_imaging}D  show the correlations between the recovered and true angles and rank orders, respectively, for the entire data set. 
%
Both the angles and the ranks are recovered with a high degree of accuracy.

@@

\subsubsection*{Zebrafish epiboly}

@@

\vspace{0cm}
\raisebox{2.5cm}{\textsf{C}}
%
\includegraphics[width=3.25cm]{zebrafish_rank_corr}
\end{minipage}
\caption{Method validation using live imaging of a zebrafish embryo. (A) Selected images from a flip book movie of zebrafish epiboly. Each frame is in an arbitrary rotational orientation, and the order of the frames has been shuffled. (B) Images from A after registration and ordering using vector diffusion maps. The real time of each frame is also indicated. (C) Correlation between the rank recovered using vector diffusion maps and the true rank.  The rank correlation coefficient is $0.9954$. The larger errors in the recovered ranks towards the beginning of the trajectory are due to the rather slow cell movement within that time window. }
\label{fig:zebrafish}
\end{figure*}

To illustrate the broad applicability of our approach, we also applied our algorithm to a time-lapse movie  of zebrafish embryogenesis. 
%
We used a publicly available flipbook movie of zebrafish embryogenesis, available at \texttt{https://zfin.org/zf\_info/movies/Zebrafish.mov} \citep{karlstrom1996flipbook}.  
%
Taken with a differential interference contrast (DIC) microscope, the movie records the first $17$~hours of zebrafish development, from a single cell stage to a 16-somite stage. 
%
We selected $120$ consecutive frames from this movie which capture 5.5~hours of epiboly (3.5--9~hours after fertilization). 
%
At the start of the time window, cells have divided 10--11 times and are accumulated in a cell mass above the yolk. 
%
The cell mass is then compressed and the animal-vegetal axis of the embryo (vertical axis in \fig~\ref{fig:zebrafish}) shortens to form a spherical embryo shape by the end of the fourth hour of development. 
%
Then, the yolk syncytial layer, which forms the boundary between the yolk and the cell mass, moves upward, forming a dome-shaped structure.  
%
During this stage, the cells rearrange to form a cell layer with uniform thickness (about four cells thick). 
%
With time, this cell layer then spreads over across the yolk and expands toward the vegetal pole. 
%
At the end of epiboly, the blastoderm completely engulfs the yolk.

As in the example of {\em Drosophila} embryo live imaging, the frames were randomly rotated and shuffled (\fig~\ref{fig:zebrafish}A). 
%
We then used vector diffusion maps to register and order the frames. 
%
The results are shown in \fig~\ref{fig:zebrafish}B. 
%
The recovered rotations and order are consistent with the expected developmental dynamics, as shown in the correlations between the recovered and true ranks (\fig~\ref{fig:zebrafish}C). 
%
Quantitatively, the rank correlation coefficient for this data set is 0.9954, and the average error in the recovered angle is $4.14^\circ$. 
%
Some errors in ordering images of the early embryo result from rather slow cell movement during the early developmental stage where cells divide and accumulate above the yolk. 
%
During epiboly, cell movement is more dynamic and the recovered ordering is more consistent with the real dynamics.

In summary, we have showed that our algorithm is not limited to a specific set of data. 
%
The algorithm works for different tissue types ({\em Drosophila} and zebrafish), and also for images taken with different imaging methods (fluorescent and phase contrast images).
%
Provided there exist significant dynamics within the data set and that the developmental trajectory is well-sampled, the developmental dynamics can be recovered.

@@

\raisebox{0.5cm}{\textsf{C}}
\includegraphics[width=12.25cm]{drosophila_fixed_images_average}
\end{minipage}
%
\hfill
%
\raisebox{3.25cm}{\textsf{D}}
\includegraphics[width=4.25cm]{drosophila_fixed_images_rank_corr}
%
\caption{Analysis of images of fixed \textit{Drosophila} embryos. {(A)} Images of \textit{ Drosophila} embryos, stained for nuclei (gray), Twi (green), and dpERK (red). Each image is of a different embryo arrested at a different developmental time and in a different rotational orientation. {(B)} Data from {\it A}, registered and ordered using vector diffusion maps. The expert rank for each image is indicated. {(C)} A representative ``developmental trajectory'' obtained from local averaging of the entire set of registered and ordered images (see \SI). {(D)} Correlation between the image ranks calculated from the vector diffusion maps algorithm and the ranks obtained from ordering by an expert. The rank correlation coefficient is $0.9716$. }
\label{fig:drosophila_fixed_images}
\end{figure*}

We have analyzed how our algorithm performs on two model data sets where all images come from a single embryo. 
%
In practice, we are interested in cases where image comes from a different embryo, and the largest source of noise in the considered data set arises from embryo-to-embryo variability.
%
To demonstrate that our methods are robust to such noise sources,
we constructed a synthetic time course data set by selecting a random image from one of five {\em Drosophila} live imaging data sets (those data sets used in \fig~\ref{fig:drosophila_live_imaging}) at each time point.
%
The resulting data set is spatially unregistered, scrambled in time, and reflects embryo-to-embryo variability. 
%
The median rank correlation coefficient when ordering such a synthetic time course using our methodology was 0.77, indicating that the algorithm can recover the temporal order even under noisy conditions.

@@

\fig~\ref{fig:drosophila_fixed_images}B shows the selected images in \fig~\ref{fig:drosophila_fixed_images}A, now registered and ordered using vector diffusion maps \citep{singer2012vector}.
%
However, because of interembryo variability, these selected images are not entirely representative of the developmental dynamics across the entire data set.
%
To highlight the developmental dynamics revealed by vector diffusion maps, we computed an average developmental trajectory across the set of registered and ordered images (see \SI). 
%
Each snapshot in the average trajectory is the (weighted) average of a group of successive images from the registered and ordered data set.
%
Averaging successive images removes some of the interembryo variability, so that sequential snapshots of this averaged trajectory, shown in \fig~\ref{fig:drosophila_fixed_images}C, serve as a summary of the stereotypic developmental dynamics.

From this average trajectory, we can now easily see the developmental progression, which is consistent with the known dynamics.
%
dpERK first appears as two lateral peaks at the ventrolateral side of the embryo; a third dpERK peak then appears at the dorsal side of the embryo.
%
During mesoderm invagination, the two ventrolateral dpERK peaks merge together, eventually forming, together with Twi, the ``omega'' shape.
%
The dorsal dpERK peak then disappears during germband extension as cells from the ventral side wrap around to the dorsal side.
%
At the end of this process, similar ``omegas'' formed by Twi and dpERK appear on the dorsal side of the embryo; these patterns are most readily seen in the last image of \fig~\ref{fig:drosophila_fixed_images}C.
%
Thus, vector diffusion maps can accomplish the tasks presented in the caricature in \fig~\ref{fig:fish}, even in the absence of information about image landmarks (e.g., fins) and without {\it a priori} knowledge of developmental features (e.g., correlation of age with body size).
%
